Title Text

### 2014-04-01 intromoa.tex
more edits
latest version, still many bib files. Awaiting your review.

MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an algebra and a calculus of 
indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
gets its names from the {\em Psi}  function which is the foundational function in the theory. The name {\em Psi} 
was inspired by Dirac's {\em Psi Vectors} and his communications with Einstein to have a mathematical way to reason
about them. These results began with the awareness that Iverson's APL\cite{iverson62}
was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.
The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
that theoretical anamolies were introduced. More on this later.

Not only did  Ken believe that APL's array algebra was the ideal basis for a programming language, he (and others) believed that
an ideal machine language supported array operations. This led to HP building an 
APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
to define an algebra of arrays based on shapes; others followed\cite{lyon,guibas}. He believed that if array operations were based on 
indexing with  shapes, all  indexing in the array expression could be  composed
and  thus could minimize the creation (and deletion) of temporary arrays,
a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
It was  Alan Perlis  who realized that these optimizations should be done at compile time\cite{perlis}, others
followed\cite{miller,budd84}. Perlis was also the first to recognize that there were too many anamolies in both APL's
algebra and environment; he recommended a functional array calculator based on the lambda 
calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
the Lambda Lalculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
functional programming language with 
arrays\cite{mul91,budd91,hm92,mul93,mul94,mul96}.

When compiler support was implemented in APL,
the main limitation was that the anomalies in the language~\cite{gerhart} 
made it difficult to optimize and impossible to verify correctness of 
programs.

MoA and the Psi Calculus, as a theory  has  also been used to describe and verify 
hardware\cite{hardy1,hardy2,fftharry}, describe the partitioning and
scheduling of array operations (and their costs)\cite{mdst93,lee95,aachen96,coffin94,ll2,mcmahon95,haleh98}, 
and the formulation of parallel and distributed processing\cite{Mul03}. 
MoA and $\psi$-Calculus, can be used to abstract complex processor memory layouts 
for tensor/array expressions, reduce these expressions first to a semantic 
normal form (Denotational Normal Form-DNF) then to an operational normal form 
(ONF). The ONF describes how to {\em build} the code using 
{\em start, stop, stride}, and {\em count}: a universal machine abstraction.
This is possible by {\em abstractly} lifting the dimension of an expression to include all aspects
of a processor, memory, communication topology. . 
%
%
%
%\bibliographystyle{plain}
%\bibliography{paper1,hpf,workshop}
\section*{Elements of the theory}
\subsection*{\label{indshp} Indexing and Shapes}

@@

\hspace{2cm}\= $u \;\cat\; v$ \hspace{1cm} \=		\=catentation of vectors $u$ and $v$ \\
\>$u$ + $v$\>		\>itemwise vector addition assuming $t_u = t_v$ \\
\>$u \times v$ \>	\>	itemwise vector multiplication \\
\>$n$ + $u$, $u$ + $n$ \>	addition of a scalar to each item of a vector \\
\>$n \times u$, $u \times n$ \>   multiplication of each item of a vector by a scalar \\
\>$\iota \; n$ \>\>  the vector of the first n integers starting from $0$ \\
%\footnote{$\iota \vec n$ is also defined and  produces an array of 
%cartesian coordinates~\cite{mul88}.}\\
\>$\pi \; v$ \> \>a scalar which is the product of the components of v \\
\>$k \;\take \;u$ \>\>	when $k \geq 0$ the vector of the first $k$ items of $u$,(called {\em take}) \\
\> \> \>and when $k<0$ the vector of the last $k$ items of $u$ \\
\>$k \;\drop \;u$ \>  \> when $k \geq 0$ the vector of $t_u - k$ last items of $u$,(called{ \em drop}) \\
\> \> \>and when $k<0$ the vector of the first $t_u - |k|$ items of $u$ \\
\>$k \;\theta \;u$ \>  \>  when $k \geq 0$  the vector of $(k  \drop  u) \cat ( k \take u ) $\\
\> \> \>and when $k<0$ the vector or $(k \take u ) \cat ( k \drop u ) $
\end{tabbing}

@@

\section*{Higher Order Operations}

>>>>

MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an algebra and a calculus of 
indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
gets its names from the {\em Psi}  function which is the foundational function in the theory. The name {\em Psi} 
was inspired by Dirac's {\em Psi Vectors} and his communications with Einstein to have a mathematical way to reason
about them. These results began with the awareness that Iverson's APL\cite{iverson62}
was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.
The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
that theoretical anamolies were introduced.

Not only did  Ken believe that APL's array algebra was the ideal basis for a programming language, he (and others) believed that
an ideal machine language supported array operations. This led to HP building an 
APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
to define an algebra of arrays based on shapes; others attempted to
put closure on his work\cite{lyon,guibas}. Abrams believed that if array operations were based on 
indexing with  shapes, all  indexing in the array expression could be  composed
and  thus could minimize the creation (and deletion) of temporary arrays,
a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
It was  Alan Perlis, and later others\cite{miller,budd84},  who realized that these optimizations should be done at compile time. Attempts at compilation led Perlis to realize that  anomalies in the language~\cite{gerhart} 
made it difficult to optimize and impossible to verify correctness of 
programs leading him to recommend a functional array calculator based on the lambda 
calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
the Lambda Lalculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
functional programming language with 
arrays\cite{mul91,budd91,hm92,mul93,mul94,mul96}.

MoA and the Psi Calculus, as a theory,  have  also been used to describe and verify 
hardware\cite{hardy1,hardy2,fftharry}, describe the partitioning and
scheduling of array operations (and their costs)\cite{mdst93,lee95,aachen96,coffin94,ll2,mcmahon95,haleh98}, 
and the formulation of parallel and distributed processing\cite{Mul03}. 
MoA and $\psi$-Calculus, can be used to abstract complex processor memory layouts 
for tensor/array expressions, reduce these expressions first to a semantic 
normal form (Denotational Normal Form-DNF) then to an Operational Normal Form 
(ONF). The ONF describes how to {\em build} the code using 
{\em start, stop, stride}, and {\em count}: a universal machine abstraction.
This is possible by {\em abstractly} lifting the dimension of an expression to include all aspects
of a processor, memory, communication topology. For example, to map a 2-d array expression to
n  processors we would lift the dimenion to 3-d. 
%
%
%
%\bibliographystyle{plain}
%\bibliography{paper1,hpf,workshop}
\subsection*{Elements of the theory}
\subsection*{\label{indshp} Indexing and Shapes}

@@

\hspace{.25cm}\= $u \;\cat\; v$ \hspace{1cm} \=		\=catentation of vectors $u$ and $v$ \\
\>$u$ + $v$\>		\>itemwise vector addition assuming $t_u = t_v$ \\
\>$u \times v$ \>	\>	itemwise vector multiplication \\
\>$n$ + $u$, $u$ + $n$ \>	addition of a scalar to each item of a vector \\
\>$n \times u$, $u \times n$ \>   multiplication of each item of a vector by a scalar \\
\>$\iota \; n$ \>\>  the vector of the first n integers starting from $0$ \\
%\footnote{$\iota \vec n$ is also defined and  produces an array of 
%cartesian coordinates~\cite{mul88}.}\\
\>$\pi \; v$ \> \>a scalar which is the product of the components of v \\
\>$k \;\take \;u$ \>\>	when $k \geq 0$ the vector of the first $k$ items of $u$,(called {\em take}) \\
\> \> \>and when $k<0$ the vector of the last $k$ items of $u$ \\
\>$k \;\drop \;u$ \>  \> when $k \geq 0$ the vector of $t_u - k$ last items of $u$,(called{ \em drop}) \\
\> \> \>and when $k<0$ the vector of the first $t_u - |k|$ items of $u$ \\
\>$k \;\theta \;u$ \>  \>  when $k \geq 0$  the vector of $(k  \drop  u) \cat ( k \take u ) $\\
\> \> \>and when $k<0$ the vector or $(k \take u ) \cat ( k \drop u ) $
\end{tabbing}

@@

\subsection*{Higher Order Operations}
